---
title: "BankingChurn.Rmd"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2025-12-23"
---

## Refernce used to setup and the XGBoost GPU in windows -> "https://stackoverflow.com/questions/77931563/how-to-use-xgboost-in-r-with-gpu"

```{r Setup, echo=TRUE, message=FALSE, warning=FALSE, }
knitr::opts_chunk$set(echo = TRUE)

force_reinstall_everything <- FALSE ## <--- change to TRUE to force a reinstall of 
                                    ## all packages listed in allowed.packages

## ADD ANY REQUIRED PACKAGES TO THE VECTOR BELOW
## Note: tidyverse includes:  ggplot2, dplyrtidyrreadrpurrrtibble, 
##                            stringrforcats, and lubridate.
allowed.packages <- c("tidyverse", 
                      "tidymodels", 
                      "recipes", 
                      "parsnip", 
                      "devtools",
                      "bonsai", 
                      "healthyR.ai", 
                      "knitr", 
                      "forcats", 
                      "DataExplorer", 
                      "visdat", 
                      "rsample", 
                      "multcomp",
                      "dplyr", 
                      "themis", 
                      "jsonlite", 
                      "data.table"
                      )


## DO NOT CHANGE ANYTHING ELSE IN THIS CODE BLOCK
for (i in seq_along(allowed.packages)){
  if(!(allowed.packages[i] %in% installed.packages()[, 1]) || force_reinstall_everything){
    install.packages(allowed.packages[i], 
                     dependencies = c("Depends", "Imports"),
                     repos = "https://cloud.r-project.org")
  }
  library(allowed.packages[i], character.only = TRUE)
}

knitr::opts_chunk$set(echo = TRUE, collapse=TRUE)
ignore <- runif(1) # To initialise the PRNG

cat("\nList of allowed packages (pre-loaded):\n ", paste(allowed.packages, collapse="\n  "))

```

```{r}

# Load it 
library(xgboost)

```

## Load the CSV file from the path "D:/VS_Codes/DataScience/Projects/Banking_Customer_Churn/r_xgb/BankChurners.csv"

```{r}
# Seed values 
seed_value = 123

# File path of the CSV file 
file_path = "D:/VS_Codes/DataScience/Projects/Banking_Customer_Churn/r_xgb/BankChurners.csv"

# Load the CSV with read csv method and store it in df 
df = read.csv(file = file_path, header = TRUE)

# Print the head to check the CSV loaded or not 
head(df)

```
## EDA and Preprocessings 

## Basic EDA to check the detailes about the dataset 
  # Glimpse the dataset 
  # Columns Names 
  # Decide the features and labels 
  # Check the propositions of the feature and plot it
```{r}

# Glimplse the df 
glimpse(df)

```
```{r}

# Get a table of all classes
table(sapply(df, class))

```


```{r}

colnames(df)

```
The data frame has 22 columns, the feature columns is the binary values which indicates the churn state of the customer. There are 6 character columns, 15 Numeric columns, and a logical columns.   

```{r}

# Summarize it as a dataframe and plot the bar graph 
feature_count.df = as.data.frame(prop.table(table(df$Attrition_Flag)) * 100)
feature_count.df

# Set the column names 
colnames(feature_count.df) = c("Attrition_Flag", "Count_percentage")

# Plot the percentage of the Attrition Flag
ggplot(feature_count.df, aes(x = Attrition_Flag, y = Count_percentage, fill = Attrition_Flag)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Count_percentage, 1)),
            vjust = -0.5) + 
  labs(title = "Feature Distribution", x = "Attrition_Flag", y = "Count_percentage") + 
  theme(panel.grid.major.x = element_blank())

```

## Factor the label column 
```{r}

# Factor the Attrition_Flag
df$Attrition_Flag = as.factor(df$Attrition_Flag)

```

The feature columns is imbalanced in the ratio of 15:58, so the feature column is under sample or over sample before the model training. 


From the columns details, Attrition_Flag is the label to predict about the churn of the bank and rest of the columns are the labels. 

## Processing pipeline with the recipe 
# Split the data set as train test split for the preprocessing and EDA 
```{r}

# Set the seed 
set.seed(seed_value)

# Split the dataset and store it in df.split using initial split method.
df.split = initial_split(df, prop = 0.8)


# Store the train data in df_train and test data in df_test 
df_train = training(df.split)
df_test = testing(df.split)


# Check train dataset 
colnames(df_train) 
```


  # Create a recipe with the train data set and store it in the variable
```{r}

# Create a recipe and store it in a recipe.df with the feature column as Attrition_Flag
recipe.df = recipe(Attrition_Flag ~ ., data = df_train)

recipe.df
```

## Check the NA values in the entire dataset and plot it with the box plot 
  # Create a dataframe to calculate the NA values in each column 
  # Plot the bar graphs for the NA data frame
```{r}

# Calculate the NA values of each column in the df and store it in na.df 
na.df = data.frame(
  col = colnames(df), 
  na_sum = colSums(is.na(df))
)

na.df
```

```{r}

# Bar plot to plot the sum of columns 
ggplot(na.df, aes(x = col, y = na_sum)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  coord_flip() +
  ylab("Sum of NA") +
  xlab("Columns with NA") +
  theme(panel.grid.major.x = element_blank())

```

From the bar graph, there is only one column with more than 10K na values, so remove the columns in the recipe. 

## Remove the NA column 
```{r}

# Get the col to remove to store it in the col_remove 
col_remove =  na.df %>%
                filter(na_sum > 0) %>%
                pull(col)

col_remove


# Update the column to removed na column in recipe as columns_to_remove 
recipe.df = recipe.df %>% 
              update_role(all_of(col_remove), new_role = "columns_to_remove") 

recipe.df
```

## Check the outliers in the dataset 
  # Using IQR to find the outliers in each columns and store it in the dataframe 
  # Plot the outliers
  
```{r}

# Create a method as count_outlier in each column using the IQR and return the dataframe 
count_outlier = function(df) {
  # Get the numeric column and store it in num_cols
  num_cols = sapply(df, is.numeric)
  
  # Loop each columns in the dataframe using sapply 
  # Calculate the outliers of the each columns 
  # Store it in the ouliter.df 
  outlier.df = data.frame(
    column = names(df)[num_cols],
    n_outliers = sapply(df[, num_cols], function(col) {
      q1  <- quantile(col, 0.25, na.rm = TRUE)
      q3  <- quantile(col, 0.75, na.rm = TRUE)
      iqr <- q3 - q1
      lower <- q1 - 1.5 * iqr
      upper <- q3 + 1.5 * iqr
      sum(col < lower | col > upper)
    })
  )
  
  # Return the dataframe 
  return(outlier.df)
}

# Invoke the method count_outlier to get the outlier and store it in outlier.df
outlier.df = count_outlier(df)

outlier.df
```
# Plot the bar graphs 
```{r}

# Plot the point graphs to plot the count of the ouliers in the each columns 
ggplot(outlier.df %>% filter(n_outliers > 0), aes(x = column, y = n_outliers)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Distribution of outlier counts (per feature)",
       x = "Outlier count per feature", y = "Number of features") +
  theme(axis.text.x = element_text(angle=45, hjust = 1))

```

# Separate the columns with the outliers and winsorized it to remove the outliers 
```{r}

# Get the columns with the outliers 
outliers_cols = outlier.df %>% 
                  filter(n_outliers > 0) %>% 
                  pull(column)

outliers_cols


recipe.df = recipe.df %>%
  # winsorize only the outlier columns
  step_hai_winsorized_truncate(
    outliers_cols,
    fraction = 0.05
  )
  

recipe.df
```

# Check the Scale of the numerical columns 
  # Get the numerical columns 
  # Check the scale using the SD method 
```{r}

# Get the numerical columns and store it in df_num
df_num = df %>% select_if(is.numeric)

# Create a dataframe with the sd range 
df_sd_range = data.frame(
                Features = colnames(df_num),            
                SD = sapply(df_num, function(x) sd(x, na.rm = TRUE)),
                Mean = sapply(df_num, function(x) mean(x, na.rm = TRUE))
                ) %>% 
                mutate( 
                  Range = log10(abs(Mean - SD) + 1e-12), 
                  )

# Plot the bar graphs to plot 
ggplot(df_sd_range, aes(x = Features, y = Range)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Distribution of Scale (per feature)",
       x = "Scale values per feature", y = "Number of features") +
  theme(axis.text.x = element_text(angle=45, hjust = 1))

```

# Normalize the values in the recipe.df 
```{r}

# Normalize the numerical predictors using srep normalize 
recipe.df = recipe.df %>% 
              step_normalize(all_numeric_predictors())

recipe.df
```

## There are 4 characters column need to processed (Card_Category, Income_Category, Marital_Status & Education_Level). 
  # Check the Unique values in the each column 
  # Try to minimize the values and combine the values with same semantic meanings. 
  # Encode the values with the recipe 
```{r}

# Create a vectors of the columns names as char_cols 
char_cols = c("Card_Category", "Income_Category", "Marital_Status", "Education_Level", "Gender")

# create a function to check the unique values in each of the columns and print the dataframe with the columns names 
check_unique = function(df ,vec){
  
  # Loop the vector 
  # Get the unique values and store it in col.unique
  # Create a dataframe with vector and unique values 
  return(tibble(
    col = vec,
    unique_values = lapply(df[vec], unique)
  ) %>%
  mutate(unique_values = map_chr(unique_values, ~ paste(.x, collapse = ", "))))
}


print(check_unique(df, char_cols))
```

# The Card_Category cannot be imputed, since those values are semantically different. The Gender has only M and F are semantically differnt. 
# The Income_Category states in different ranges can be converted into Less than 40K, more than 120K, & Average Between 40K to 120K. 
# The Marital_Status classify into 4 categories with 3 having the same semantic values, they are merged into 2 categories Married and Single.  
# The Education_Level classified into 7 categories with 5 having the same semantic meaning, they are merged into 2 categories as Educated and Uneducated. 
```{r}

recipe.df = recipe.df %>% 
              # mark unseen categories when new data arrives
              step_novel(Income_Category) %>%
              # Merge the Income_Category into Less than 40K, more than 120K, & Average Between 40K to 120K
              step_mutate(Income_Category = fct_collapse(
                Income_Category, 
                "Less than 40K" = "Less than $40K",
                "More than 120K" = "$120K +",
                "Average Between 40K to 120K" = c(
                  "$40K - $60K",
                  "$60K - $80K",
                  "$80K - $120K", 
                  "Unknown"
                )
              )
            ) %>%
            # mark unseen categories when new data arrives
            step_novel(Marital_Status) %>%
            # Merge the Marital_Status into Married and Single
            step_mutate(
              Marital_Status = fct_collapse(
                Marital_Status,
                Married = "Married",
                Single  = intersect(
                  levels(Marital_Status),
                  c("Single", "Unknown", "Divorced")
                )
              )
            ) %>%
            # mark unseen categories when new data arrives
            step_novel(Education_Level) %>%
            # Merge the Education_Level into Educated and Uneducated
            step_mutate(
              Education_Level = fct_collapse(
                Education_Level, 
                  Uneducated = c(
                    "Uneducated", 
                    "Unknown"
                    ),
                  Educated = c(
                    "High School", 
                    "Graduate", 
                    "College", 
                    "Post-Graduate", 
                    "Doctorate"
                  )
                )
              )
            
recipe.df            

```
# Encode all the above categorial variables with one-hot encoding. 
```{r}

# Categorical values 
char_cols

recipe.df = recipe.df %>% 
              # One-hot encode using dummy method for all the character columns 
              step_dummy(
                char_cols, 
                one_hot = TRUE 
              )

recipe.df

```
## Find the correlation and causion between the variables to check the variables relations.
    # Plot the correlations and plot it. 
```{r}

# Find all the numerical columns 
df_num

# Find the correlation and store it in cor.mat 
cor.mat = cor(df_num, use = "pairwise.complete.obs")

cor.mat

# convert to tidy format using pivot_longer
corr_long = cor.mat %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>% 
  pivot_longer(
    cols = -var1,
    names_to = "var2",
    values_to = "correlation"
  )

# Plot the correlation between the variables as heat map 
ggplot(corr_long, aes(var1, var2, fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "steelblue",
    high = "pink",
    mid = "#90EE90",
    midpoint = 0
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "Correlation")

```

## The feature column is so unbalanced in the ratio of 15:85. To balance this, use the oversampleing SMOTE techinique. 
```{r}

recipe.df = recipe.df %>% 
              # Oversample the data using SMOTE technique 
              step_smote(Attrition_Flag)

recipe.df
```
## Basic EDA and preprocessing are done with handled the missing values, checked the outliers in the numerical features and handled it, check the scales and normalized it, check the individual categorial values, handled the semantic relations by featured and encoded it, handled the imbalance by using SMOT.

## Prepare the recipe 
```{r}

# Prepare the recipe with prep() method and store it in recipe.df_prep
recipe.df_prep = prep(recipe.df)
                    

summary(recipe.df_prep)
```

## Bake it on the data 
```{r}

# Bake the prepared recipe with bake() and store it in recipe.df_bake with the train data 
recipe.df_bake = bake(recipe.df_prep, df_train)

summary(recipe.df_bake)

```
## Modelling and Validations 

## K-fold cross validation 
  # Split the train data into k-fold to validate and train the model 
```{r}

# set the seed 
set.seed(seed_value)

# Create the k-fold using vfold_cv() and store it in k.fold 
k.fold = vfold_cv(df_train, v = 5, strata = "Attrition_Flag")

summary(k.fold)

```
## Model Architecture 
 # Create a XGBoost Model Architecture 
 # Create and fit the workflow 
 # Create an hyperparameter and tune it 
 # Collect the metrics 
 # Get the best parameter 
 # Fit the parameter with the model and create a finalized model 

## Create an architecture for the XGBoost algorithm
```{r}

# Create the XGBoost model and store it in xgboost.model 
xgboost.model = boost_tree(
  # Number of trees 
  trees = tune(), 
  # learning rate 
  learn_rate = tune(),
  # depth of tree
  tree_depth = tune(), 
) %>% 
  # Set the mode as classification
  set_mode("classification") %>%
  # Set the engine as xgboost
  set_engine(
    "xgboost",
    tree_method = "hist", 
    device = "cuda",
    max_bin     = 64,
    nthread     = 10
  )

summary(xgboost.model)  

sessionInfo()
```
## Create a workflow for the model and set the recipe and model to it
```{r}

# Create a workflow using workflow() and assign it to model.workflow 
model.workflow = workflow() %>% 
                  # Add the recipe 
                  add_recipe(recipe.df) %>%
                  # Add the model 
                  add_model(xgboost.model)

summary(model.workflow)
```

## Create a grid parameter for the Bayesian SearchCV 
```{r}

# Create a grid using extract_parameter_set_dials() and store it in model.grid 
model.grid = extract_parameter_set_dials(model.workflow) %>% 
              # update the tune grid parameters 
              update(
                # Number of trees 
                trees = trees(range = c(100, 500)),
                # learning rate
                learn_rate = learn_rate(range = c(0.0001, 0.1)), 
                # Tree depth 
                tree_depth = tree_depth(range = c(3, 8))
              )

summary(model.grid)
```

## Create a Bayesian search CV and tune it 
```{r}

# Set the seed 
set.seed(seed_value)

# Create a inital grid and store it in inital.grid
initial.grid = tune_grid(
    # workflow 
    object = model.workflow, 
    # folds 
    resamples = k.fold, 
    # Number of grids 
    grid = 10,
    # param 
    param_info = model.grid,
    # Metrics 
    metrics = metric_set(roc_auc, ppv, npv, accuracy)
  )

summary(initial.grid)

# Create a Bayesian Grid CV and store it in bayesian.cv
bayesian.cv = tune_bayes(
  # workflow
  model.workflow, 
  # set the fold 
  resamples = k.fold, 
  # parameters to tune  
  param_info = model.grid, 
  # tuned grid 
  initial = initial.grid,
  # Number of iterations 
  iter = 100, 
  # metrics 
  metrics = metric_set(roc_auc, ppv, npv, accuracy)
)

```

## Show the metrics 
```{r}
# Collect the metrics 
collect_metrics(bayesian.cv)

```


## Select the best parameter from the cv 
```{r}

# Get the best parameter from the for the f_meas and store it in best.parameters 
best.parameters = bayesian.cv %>% 
                    select_best(metric = "roc_auc")  

best.parameters
```


## Get the metrics of the best parameter
```{r}

# Extract the metrics of the best parameter and store it best.parameters_metrics
best.parameters_metrics = best.parameters %>%
  filter(.config == best.parameters$.config)

best.parameters_metrics

```


## Fit the best parameter to finalize the workflow model 
```{r}

# Fit the best parameter to get the best model and store it in final_model
final_model = model.workflow %>%
        finalize_workflow(best.parameters_metrics)

final_model

```

## Model Testing and evaluations 
  # last fit the model and get the scores of the model 
  # Plot the confusion matrix for the model valuation 
  # Plot the metrics
  # Extract the best workflow and save it 
  
# Last fit the model and obtain the score of the final model
```{r}

# Fit the best model with the trained data to get the best model and store it in final_fitted_model_dt
final_fitted_model_dt = final_model %>%
                last_fit(df.split, 
                         metrics = metric_set(accuracy, bal_accuracy, roc_auc, ppv, npv, f_meas, sens, spec, mcc))


# Collect the metrics 
collected_metrics = collect_metrics(final_fitted_model_dt)

collected_metrics
```

# Plot the metrics of the model 
```{r}

# Create a dataframe to store the metrics and estimates to plot it. Store it in metrics.df 
metrics.df = data.frame(
  metrics = collected_metrics$.metric, 
  values = collected_metrics$.estimate
)

metrics.df

```

```{r}

# Plot the metrics 
metrics.df %>%
  ggplot(aes(x = reorder(metrics, values), y = values)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Model Performance Metrics",
    x = "Metrics",
    y = "Scores"
  ) + 
  theme_minimal()

```

# Plot the confusion matrics of the model evaluation data 
```{r}

# Collect the prediction data from the last fit model and store it in pred.data 
pred.data = final_fitted_model_dt %>% 
              collect_predictions()

# Create a confusion matrix and store the values in confusion.matrix
confusion.matrix = pred.data %>% 
                    conf_mat(
                      truth = Attrition_Flag, 
                      estimate = .pred_class
                      )

summary(confusion.matrix)

# auto plot the confusion matrix 
autoplot(confusion.matrix, type = "heatmap") +
  labs(title = "Confusion Matrix")

```


# Extract the workflow and save the workflow 
```{r}

# Extract the workflow using extract_workflow() and store it in finalized_model
finalized_model = extract_workflow(final_fitted_model_dt)

finalized_model

# Save the model workflow as the rds file 
model_file.path = "D:/VS_Codes/DataScience/Projects/Banking_Customer_Churn/r_xgb"
saveRDS(finalized_model, file = file.path(model_file.path, "final_model.rds"))
```